---
title: "05_models"
author: "Lauren Howard and Kara McCormack"
date: "4/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(INLA)
library(glmmTMB)
library(tidyverse)
library(kableExtra)
library(here)
library(cowplot)
library(ggpubr)
options(scipen = 999)
```

Goal: 
  - Run GLMM models and INLA models and compare model fit
    * A. Cases, Period 1
    * B. Cases, Period 2
    * C. Mortality, Period 2
  - D. Generate Coefficient tables

Input:
  - weekly.p1.rds (weekly counts for P1 with latent classes and covariates merged in)
  - weekly.p2.rds (weekly counts for P2 with latent classes and covariates merged in)
  - nyc_map_modzcta.adj (INLA spatial file)

Output: 
  - p1_cases_weekly (formatted fixed effects for INLA model for P1 cases)
  - p2_cases_weekly (formatted fixed effects for INLA model for P2 cases)
  - p2_deaths_weekly (formatted fixed effects for INLA model for P2 deaths)

# Read in data
```{r readdata}
# set directory
dir <- here('output')

# load modeling data created in 03_data_merge.Rmd:
load('output/data.RData')

# read in adjacency file
# to do: create this in data mgmt
gr <- inla.read.graph(filename = "input/nyc_map_modzcta.adj")

# prepare data sets for analysis 
for (dataset in c('weekly.p1', 'weekly.p2')) {
  # read in covid dat
  w <- eval(parse(text = dataset))
  
  # create unique zctas
  z <- unique(w$modzcta)
  idarea <- 1:length(z)
  
  z_area <- data.frame(modzcta = z, idarea = idarea)
  
  # merge everything: now has counters for area and time
  w2 <- w %>%
    left_join(z_area, by = "modzcta") 
    # relabel comorbidity classes 
    # mutate(com_class = factor(com_class,
    #                           levels = c("3 (Low)",
    #                                      "2 (Medium)",
    #                                      "1 (High)")), 
    #        ses_class = factor(ses_class, 
    #                           levels = c("Low Dis., High Adv.", 
    #                                      "Low Dis., Low Adv.", 
    #                                      "High Dis., High Adv.", 
    #                                      "High Dis., Low Adv.")))

  w2 <- w2 %>% rename('idtime' = 'seq.w')
  w2$idtime_dup <- w2$idtime
  
  assign(dataset, w2)
}
```

What are the raw proportions of covariates for each latent variable class?
```{r}
one_dayp1 <- daily.p1 %>%
  filter(date == "2020-06-08")

table(one_dayp1$adv_level)

one_dayp1 %>%
  ggplot(aes(x = age_over_65_prop)) +
  geom_histogram(bins = 45) +
  facet_grid(~adv_level)

barplot_dat.adv <- daily.p1 %>%
  filter(date == "2020-06-08") %>%
  group_by(adv_level) %>%
  summarize(mean_ageover65 = mean(age_over_65_prop), 
            mean_nhb = mean(pop_non_hispanic_black_prop), 
            mean_hisp = mean(hispanic_prop)) %>%
  pivot_longer(cols = mean_ageover65:mean_hisp, 
               names_to = "covariate", 
               values_to = "value")

barplot_dat.adv %>%
  ggplot(aes(x = covariate, y = value, fill = adv_level)) +
  geom_bar(stat = "identity", 
           position = "dodge")

# logistic model
logistic.dat <- one_dayp1 %>%
  mutate(adv_level = ifelse(adv_level == "High", 1, 0))
model <- glm(adv_level ~ age_over_65_prop + pop_non_hispanic_black_prop + hispanic_prop, 
             family = binomial(link = 'logit'), 
             data = logistic.dat)
summary(model)
```

```{r}
# barplots of raw proportions of covariates in each SES grouping
ses_covar.bar <- one_dayp1 %>%
  select(modzcta, ses_class, age_over_65_prop, pop_non_hispanic_black_prop, hispanic_prop) %>%
  group_by(ses_class) %>%
  summarize(`Age Over 65` = mean(age_over_65_prop), 
            `Non-Hispanic Black` = mean(pop_non_hispanic_black_prop), 
            `Hispanic` = mean(hispanic_prop)) %>%
  pivot_longer(cols = `Age Over 65`:`Hispanic`, 
               names_to = "covar", 
               values_to = "value") %>%
  ggplot(aes(x = covar, y = value)) +
  geom_bar(stat = "identity", width = .8) +
  facet_grid(~ses_class) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text.x = element_text(size = 6)) +
  xlab("Covariates to SES LCM") +
  ylab("Mean Proportion")
ses_covar.bar
```
```{r}
# raw proportions for COM groupings
com_covar.bar <- one_dayp1 %>%
  select(modzcta, com_class, age_over_65_prop, pop_non_hispanic_black_prop, hispanic_prop) %>%
  group_by(com_class) %>%
  summarize(`Age Over 65` = mean(age_over_65_prop), 
            `Non-Hispanic Black` = mean(pop_non_hispanic_black_prop), 
            `Hispanic` = mean(hispanic_prop)) %>%
  pivot_longer(cols = `Age Over 65`:`Hispanic`, 
               names_to = "covar", 
               values_to = "value") %>%
  ggplot(aes(x = covar, y = value)) +
  geom_bar(stat = "identity", width = .8) +
  facet_grid(~com_class) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  xlab("Covariates to Comorbidity LCM") +
  ylab("Mean Proportion")
com_covar.bar

# hello
```

```{r}
bar.lcm.covar <- ggarrange(ses_covar.bar, com_covar.bar, 
          labels = c("A", "B"),
          ncol = 2, nrow = 1)
ggsave(filename = "figures/S2bar_lcm.png",
       plot = bar.lcm.covar,
       width = 8, height = 3, units = "in")
```


How many cases and deaths at end of each period? 
```{r}
daily.p1 %>%
  filter(date == "2020-06-08") %>%
  summarize(cases = sum(case_count))

daily.p2 %>%
  filter(date == "2021-05-10") %>%
  summarize(cases = sum(case_count))

daily.p2 %>%
  filter(date == "2021-05-10") %>%
  summarize(cases = sum(death_count))
```

How many modzctas in each class? 
```{r}
last_day <- daily.p1 %>%
  filter(date == "2020-06-08")
table(last_day$ses_class)
table(last_day$com_class)
```


While the Moran results already inform that a BYM spatial structure is needed (hence the use of INLA), for a further check, let's assess model fit using glmmTMB. 

# glmmTMB models
```{r glmmtmb_models}
# Period 1
# Negative binomial model
glmmtmb_cases_p1 <- glmmTMB(case_weekly1 ~ 
                           scale(male_prop) +
                           idtime + 
                           com_class*idtime + 
                           ses_class*idtime +
                           scale(age_over_65_prop) +
                           scale(pop_non_hispanic_black_prop) +
                           scale(hispanic_prop) +
                           scale(age_over_65_prop) +
                           com_class +
                           ses_class +
                           offset(log(pop_denom)) + 
                           (idtime|modzcta), 
                         data = weekly.p1,
                         family = nbinom2)

# Period 2
# Negative binomial model
glmmtmb_cases_p2 <- glmmTMB(case_weekly1 ~ 
                           scale(male_prop) +
                           idtime + 
                           com_class*idtime + 
                           ses_class*idtime +
                           scale(age_over_65_prop) +
                           scale(pop_non_hispanic_black_prop) +
                           scale(hispanic_prop) +
                           scale(age_over_65_prop) +
                           com_class +
                           ses_class +
                           offset(log(pop_denom)) + 
                           (idtime|modzcta), 
                         data = weekly.p2,
                         family = nbinom2)


# Period 2
# Zero inflated negative binomial model
glmmtmb_deaths_p2 <- glmmTMB(death_weekly1 ~ 
                           scale(male_prop) +
                           idtime + 
                           com_class*idtime + 
                           ses_class*idtime +
                           scale(age_over_65_prop) +
                           scale(pop_non_hispanic_black_prop) +
                           scale(hispanic_prop) +
                           scale(age_over_65_prop) +
                           com_class +
                           ses_class +
                           offset(log(pop_denom)) + 
                           (idtime|modzcta), 
                         data = weekly.p2,
                         family = nbinom2,  ziformula  = ~ 1)
```


# INLA models
Incorporates spatial correlation
```{r inla_models message = F}
# Weekly INLA negative binomial model formula with default priors and 
# random slope for time
inla_cases_formula <- case_weekly1 ~ 
  scale(male_prop) +
  idtime + 
  com_class*idtime + 
  ses_class*idtime +
  scale(age_over_65_prop) +
  scale(pop_non_hispanic_black_prop) +
  scale(hispanic_prop) +
  scale(age_over_65_prop) +
  com_class +
  ses_class +
  f(idarea, model = "bym", graph = gr) + # default priors
  f(idtime_dup, model = "iid") # random slope for time

# Period 1 CASES
# Negative binomial model with default priors
inla_cases_p1 <- inla(inla_cases_formula,
                    family = "nbinomial",
                    data = weekly.p1,
                    control.predictor = list(compute = TRUE),
                    control.compute = list(dic = TRUE),
                    verbose=TRUE,  offset = log(pop_denom))

# Period 2 CASES
# Negative binomial model with default priors
inla_cases_p2 <- inla(inla_cases_formula,
                    family = "nbinomial",
                    data = weekly.p2,
                    control.predictor = list(compute = TRUE),
                    control.compute = list(dic = TRUE),
                    verbose=TRUE,  offset = log(pop_denom))

# Weekly INLA zero-inflated negative binomial model with default priors
# and random slope for time
weekly_deaths_formula <- death_weekly1 ~ 
  scale(male_prop) +
  idtime + 
  com_class*idtime + 
  ses_class*idtime +
  scale(age_over_65_prop) +
  scale(pop_non_hispanic_black_prop) +
  scale(hispanic_prop) +
  scale(age_over_65_prop) +
  com_class +
  ses_class +
  f(idarea, model = "bym", graph = gr) + # default priors
  f(idtime_dup, model = "iid") # random slope for time

# Period 2 DEATHS
inla_deaths_p2 <- inla(weekly_deaths_formula,
                 family = "zeroinflatednbinomial1",
                 data = weekly.p2,
                 control.predictor = list(compute = TRUE),
                 control.compute = list(dic = TRUE),
                 inla.setOption(mkl=TRUE), 
                 verbose = T,
                 offset = log(pop_denom))
```

# Compare fit of glmmTMB and INLA models

```{r fitplotfun}
### Function to show fitted vs. observed values for glmmTMB and INLA models
fitplot <- function(data.in, outvar, glmmtmb_mod, inla_mod, outlab, period, xmax, ymax){
  # observed cases
  obs_cases_p1 <- data.in %>% group_by(.[,outvar]) %>% summarize(n_obs = n()) %>% rename('count' = outvar)
  
  # extract expected counts from glmmTMB model
  data.in$fitted_glmmtmb  <- round(predict(object = glmmtmb_mod, type = 'conditional', newdata =data.in))
  fitted_glmmtmb <- data.in %>% 
    group_by(fitted_glmmtmb) %>% 
    summarize(value = n()) %>% 
    rename('count' = 'fitted_glmmtmb')
  
  # extract expected counts from INLA model
  data.in$fitted_inla <- round(inla_mod$summary.fitted.values$mean)
  fitted_inla <- data.in %>% 
    group_by(fitted_inla) %>% 
    summarize(value = n()) %>% 
    rename('count' = 'fitted_inla')
  
  # glmmTMB plot
  p1 <- ggplot() +
    geom_bar(data= obs_cases_p1, aes(x=count, y = sqrt(n_obs)), stat="identity", fill = 'blue3', color = 'blue3') +
    geom_line(data= fitted_glmmtmb, aes(x=count, y = sqrt(value)), size = 1, position=position_jitter(w=1, h=0)) + 
    ylab('sqrt(Frequency)') + 
    xlim(0, xmax) +
    ylim(0, ymax) +
    ggtitle(paste0('Fitted vs Observed Plot: \nglmmTMB Model for Weekly ', outlab, 's in ' , period)) + 
    xlab(paste(outlab, ' Count'))
  
  # INLA plot 
  p2 <- ggplot() +
    geom_bar(data= obs_cases_p1, aes(x=count, y = sqrt(n_obs)), stat="identity", fill = 'blue3', color = 'blue3') +
    geom_line(data= fitted_inla, aes(x=count, y = sqrt(value)), size = 1, position=position_jitter(w=1, h=0)) + 
    ylab('sqrt(Frequency)') + 
    xlim(0, xmax) +
    ylim(0, ymax) +
    ggtitle(paste0('Fitted vs Observed Plot: \nINLA Model for Weekly ', outlab, 's in ' , period)) + 
    xlab(paste(outlab, ' Count'))
  
  plot_grid(p1, p2, align="h", ncol=2) 
}
```

## Period 1 Cases
```{r p1casefit}
### Period 1 cases
fitplot(weekly.p1, "case_weekly1", 
        glmmtmb_cases_p1, 
        inla_cases_p1, 
        "Case", "P1", 1000, 5)
```


## Period 2 Cases
```{r p2casefit}
### Period 2 cases
fitplot(weekly.p2, "case_weekly1", 
        glmmtmb_cases_p2, 
        inla_cases_p2, 
        "Case", "P2", 700, 8)
```

## Period 2 Deaths
```{r p2deathfit}
### Period 2 cases
fitplot(weekly.p2, "death_weekly1", 
        glmmtmb_deaths_p2, 
        inla_deaths_p2, 
        "Death", "P2", 50, 8)
```
While P1 cases shows similar results for glmmTMB and INLA, INLA demonstrates improvements in fit for P2 cases and P2 mortality. Based on the fit plots, we will use INLA models.


```{r format_models_fun}
# Function to fix models for display purposes 
format_model <- function(data.in){
  data.in %>% 
    mutate(var = rownames(data.in)) %>%
    filter(var != '(Intercept)') %>% 
    mutate(var = case_when(var == 'idtime' ~ 'Time',
                           var == 'com_classMedium'  ~ 'Medium comorbidity',
                           var == 'com_classHigh' ~ 'High comorbidity',
                           var == 'ses_classLow Dis., Low Adv.' ~ 'Low Disadv., Low Adv.',
                           var == 'ses_classHigh Dis., High Adv.' ~ 'High Disadv., High Adv.',
                           var == 'ses_classHigh Dis., Low Adv.' ~ 'High Disadv., Low Adv.',
                           var == 'scale(age_over_65_prop)' ~ 'Over 65',
                           var == 'scale(pop_non_hispanic_black_prop)' ~ 'Non-Hispanic Black',
                           var == 'scale(hispanic_prop)' ~ 'Hispanic',
                           var == 'idtime:com_classMedium' ~ 'Time:Medium comorbidity',
                           var == 'idtime:com_classHigh' ~ 'Time:High comorbidity',
                           var == 'idtime:ses_classLow Dis., Low Adv.' ~ 'Time:Low Disadv., Low Adv.',
                           var == 'idtime:ses_classHigh Dis., High Adv.' ~ 'Time:High Disadv., High Adv.',
                           var == 'idtime:ses_classHigh Dis., Low Adv.' ~ 'Time:High Disadv., Low Adv.',
                           var == 'scale(male_prop)' ~ 'Male'),
           expmean=sprintf("%.3f", round(exp(mean), 3)),
           CrI = paste0("[", 
                        sprintf("%.3f", round(exp(mean-1.96*sd), 3)), 
                        ", ", 
                        sprintf("%.3f", round(exp(mean+1.96*sd), 3)), 
                        "]")) %>%
    select(var, expmean, CrI) 
}
```

```{r format_models}
# extract fixed effects from models
# format model output to prepare for table

# Period 1 Weekly Cases
p1_cases_weekly <- format_model(inla_cases_p1$summary.fixed) 

# Period 2 Weekly Cases
p2_cases_weekly <- format_model(inla_cases_p2$summary.fixed) 

# Period 2 Weekly Deaths
p2_deaths_weekly <- format_model(inla_deaths_p2$summary.fixed) 
```


## Export model output
```{r exportmods}
save(p1_cases_weekly, p2_cases_weekly, p2_deaths_weekly, file="results/final.models.Rdata")
```


